Chapter 16: Particle-laden gravity currents: the lock-release slumping regime at the laboratory scale

### 2023-07-21 paper/main.tex
SUBMISSION

\documentclass[twocolumn]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{natbib}
\usepackage{graphicx, caption}
\usepackage{authblk}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{amsmath}

% \usepackage{moreverb,savetrees}

@@

\graphicspath{{./figures/}{./}}

\title{Title}
\author[1]{PALAGRAM}

\affil[1]{Affiliations}

\date{}

\newenvironment{Figure}
  {\par\medskip\noindent\minipage{\linewidth}}
  {\endminipage\par\medskip}

@@

\twocolumn[
	\maketitle

	\begin{abstract}
		Particle-laden gravity currents (PLGCs) are driven by a mass difference between a heavy fluid-particle mixture and a lighter ambient liquid. We focus our attention on finite volume current, generated here by lock-release device over a controlled constant slope bottom at the laboratory scale. The main objective of the present chapter is to provide a comprehensive flow regime associated with the relevant dimensionless parameters which characterize the current at the macroscopic scale, i.e. average over the current depth.
	\end{abstract}
	\vskip 0.7cm
]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Figure 1
\begin{figure*}[h]
	\centering
	\includegraphics{figure1.pdf}
	\caption{\textbf{Experimental set-ups used in this study.} (a--c) Sketches of the three experimental set-ups. Below are shown snapshots of corresponding experimental images. (d--h) Set-up 1, $\alpha = 7^\circ$, $\phi \sim 3~\%$, $\mathcal{S}t=$ (i--k) Set-up 2, $\alpha = $, $\phi \sim $, $\mathcal{S}t=$ (l--p) Set-up 3, $\alpha = $, $\phi \sim $, $\mathcal{S}t=$.}
	\label{fig:fig1}
\end{figure*}

\section{Introduction}

\subsection{Control parameters}

Particle-laden gravity currents (PLGCs) are driven by a mass difference between a heavy fluid-particle mixture and a lighter ambient liquid. A major difference between compositional currents and PLGCs is therefore the existence of typical length and time scales associated with the particle dynamics. Particularly, the own settling velocity of a heavy particle only allows the existence of these gravity currents on finite time and length scales. Nevertheless, prediction of these finite scales is difficult as it shall strongly depends on the interaction between particle dynamics and the local turbulence of the carrying fluid and/or ambient fluid at the upper interface, which modifies entrainment and associated dissipation, and in turn can delay or enhance settling. Accordingly, their dynamics are controlled by several dimensionless parameters leading to a wide variety of flow regimes. In particular, beyond the slope $\alpha$, the dimensionless density difference between the current and the ambient $At$ and the ratio between available inertia and dissipation, say a Reynolds number $Re$, as usually introduced as a control parameter for compositional currents, the initial solid fraction of particles $\phi$ and a dimensionless settling velocity St have now to be considered as potentially affecting flow regimes. It shall be noted that in some specific situations, some of these dimensionless parameters could be linked as, for instance, $\phi$ and $At$ for heavy particles in a single-phase liquid for the carrying and ambient fluids

\subsection{Literature}

\begin{itemize}
	\item literature slope
	\item literature settling
	\item literature volume fraction
	\item other effects: lock geometry, non-boussinesq, entrainment, erosion/deposition $\rightarrow$ ignored here
\end{itemize}

\section{Dimensional analysis at the macroscopic scale of the current}
\label{sec:dimensionlessmap}

Here, the dimensionless parameters are based on a characteristic length $L_0$ and a characteristic velocity $U_0=\sqrt{g' H_0}$ where $L_0$ and $H_0$ are the horizontal length and depth of the initial container, respectively, $g'$ is the modified gravity $r g \cos{\alpha}$ and $r = (\rho_m - \rho_a)/\rho_a$ with $\rho_m$ and $\rho_a$ the current density and ambient fluid density respectively. Note that $\rho_m = \phi \rho_p+(1-\phi)\rho_f$ with $\rho_p$ the particles density and $\rho_f$ the fluid phase density initially in the current. In the set of experiments presented here $\rho_f\le \rho_a$. The velocity scale $U_0$ corresponds to an idealized transfer of initial potential energy to kinetic energy along the slope, i.e. a collapse of the initial column.

Accordingly,
\begin{equation}
	\displaystyle a =\frac{H_0}{L_0},
\end{equation}
\begin{equation}
	\displaystyle Re = \frac{U_0 H_0\rho_m}{\eta_f},
\end{equation}
\begin{equation}
	\displaystyle At = \frac{\rho_m-\rho_a}{\rho_a},
\end{equation}
\begin{equation}
	\displaystyle St = \frac{V_s}{U_0},
\end{equation}
with $\eta_f$ the interstitial fluid viscosity and $V_s$ the Stokes settling velocity in the interstitial fluid phase labelled $f$.

Then, the dynamics of the current are characterized by several dimensionless variables. The most common one to be analyzed in gravity-driven avalanches is the dimensionless front velocity, also referred to as the Froude number,
\begin{equation}
	\displaystyle \mathcal{F}r =\frac{U_c}{U_0},
\end{equation}
where $U_c$ is the front velocity of the current. Note that $U_c$ is time-dependent, and so is $\mathcal{F}r$, i.e. that $\mathcal{F}r$ as defined here is not a control parameter. (The Froude number which would be defined as a control parameter for a lock-release configuration would be $Fr\equiv 1$, and is thus not relevant to predict the flow dynamics.)

\section{Methods}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Figure 2
\begin{figure}
	\centering
	% \includegraphics{Figures/Figure_lit_plat/Figure_lit_plat.pdf}
	\includegraphics{figure2.pdf}
	\caption{\textbf{Explored parameter space.} Each symbol correspond to a single experimental run.}
	\label{fig:fig2}
\end{figure}

\subsection{Experimental setups}

Even if the dynamics of PLGCs is associated with several complex processes, a general consequence is observed on the front dynamics, often referred to as the $\mathcal{F}r$ number, and the front shape. Then prediction of the different flow regimes associated with PLGCs shall be entirely included into these observables. In order to characterize the different flow regimes for PLGCs induced by a finite volume of material initially at “rest”, 3 dam-break-type devices have been used to cover a wide range of control parameters, as $\alpha \in [0^o,30^o]$, $\phi\in[0.1\%,20\%]$, $At\in[???]$, $Re\in[???]$, $St\in[???]$, $a\in??$, as specifically defined in section \ref{sec:dimensionlessmap}. These experimental devices are supplemented with Euler-Euler numerical simulations to confirm the relevance of such flow regime description to characterize the dynamics of PLGCs.

@@

Specifications of IMFT set-up

\paragraph{Set-up 2}

Specifications of LEGI set-up

@@

Specifications of LEMTA set-up

\subsection{Datasets}

On these 3 devices, a total of $N$ experiments are presented here, complemented by $N$ numerical simulations using SedFoam (see description below). They are classified into 4 different datasets, described below.

@@

Description Dataset LEGI

@@

Description Dataset IMFT

@@

Description Dataset LEGI/IMFT

@@

Description Dataset LEMTA

\paragraph{SedFoam}

Description Dataset SedFoam

\subsection{Fitting procedure and front position analysis}

For each experimental/numerical run, we extract the current front position as a function of time (see Fig. \ref{fig:fig3}a). A few selected runs are shown on figure~\ref{fig:fig3}b in non-dimensional time coordinates. After a short transient acceleration phase, most currents exhibit at first a constant velocity corresponding to the so-called \emph{slumping} regime. After this, the velocity can decrease, either because of settling induced buoyancy loss, or because the current enters the \emph{inertial} regime.

It can be qualitatively seen that increasing the bottom slope increases the slumping velocity, while increasing the Stokes number shortens the duration of the constant-velocity regime.
%
To properly quantify that, we fit all front propagation curves (except Dataset 4) with the following equation:

\begin{equation}
	\label{eq:fit_eq}
	\frac{x_{\rm f}}{L_{0}} = x_{i} + \mathcal{F}r \frac{t}{t_{0}} - \frac{\tilde{\lambda}}{2} \left(\frac{t}{t_{0}}\right)^{2},
\end{equation}
where $\mathcal{F}r$ corresponds now to the non-dimensional current velocity during the slumping regime, and $\tilde{\lambda}$ to the dissipation. The fit is performed between $t/t_{0} = 5$, to avoid the initial transient phase, and $t/t_{0} = 30$, the transition time from the slumping to the inertial regime~\citep{refs}. Note that $x_{i}$ is kept constant equal to 0 for Dataset 1, 2 and 3, but had to be left free to adjust for the SedFoam simulations \textcolor{red}{comment this}. For runs with strong dissipation ($\tilde{\lambda} > 2 {\times} 10^{-2}$), we add a quartic term to \eqref{eq:fit_eq} to improve the fit quality and obtain a better estimation of $\mathcal{F}r$ and $\tilde{\lambda}$. We checked that our results does not depend on the presence of this quartic term, and that the linear and quadratic ones remain much larger.

For Dataset 4, the time resolution as well as the experiment duration does not allow to extract the dissipation. Therefore, we simply fit the whole available points with a linear function.

>>>>

\documentclass[twocolumn]{article}
\usepackage[utf8]{inputenc}
\usepackage{natbib}
\usepackage{graphicx, caption}
\usepackage{authblk}
\usepackage{textcomp}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{amsmath}
\usepackage{cuted}

@@

\newcommand*{\cyril}{\textcolor{ForestGreen}}
\newcommand*{\marie}{\textcolor{Cyan}}
\newcommand*{\julien}{\textcolor{Brown}}
\newcommand*{\yvan}{\textcolor{Blue}}

\definecolor{jeanc}{RGB}{253,129,43}
\newcommand*{\jean}{\textcolor{jeanc}}

\graphicspath{{./figures}{./}}

\title{Chapter 16: Particle-laden gravity currents: the lock-release slumping regime at the laboratory scale}

% \title{Chapter 16: Lock-release particle-laden gravity currents at the laboratory scale}

% \title{Macroscopic description of particle-laden gravity currents: overall perspectives}

% \title{Macroscopic description of lock-release particle-laden gravity currents}

\author[1]{Gadal C.}
\author[2]{Schneider J.}
\author[3]{Bonamy C.}
\author[3]{Chauchat J.}
\author[2]{Dossmann Y.}
\author[2]{Kiesgen de Richter S.}
\author[1]{Mercier M. J.}
\author[4]{Naaim-Bouvet F.}
\author[3]{Rastello M.}
\author[1]{Lacaze L.}

\affil[1]{Institut de Mécanique des Fluides de Toulouse (IMFT), Université de Toulouse, CNRS, Toulouse, France}
\affil[2]{Laboratoire Énergies et Mécanique Théorique
et Appliquée (LEMTA), Université de Lorraine, CNRS, 54500, Nancy, France}
\affil[3]{Univ. Grenoble Alpes, CNRS, Grenoble INP, LEGI, 38000 Grenoble, France}
\affil[4]{Univ. Grenoble Alpes, INRAE, CNRS, IRD, Grenoble INP, IGE, 38000 Grenoble, France}

\date{{\it\color{red}Deadline: May 31$^{st}$ - repoussé (mi juillet ou EuroMech?)}}

@@

\twocolumn[

	\maketitle

	\begin{abstract}

		Particle-laden gravity currents (PLGCs) are driven by the mass difference between a heavy fluid-particle mixture and a lighter ambient liquid. They often occur in natural and industrial situations, among which a typical situation is the release of a finite volume. Here, we focus on such `dam-break' situations, which are studied using lock-release devices at the laboratory scale.
		%
		The objective of the present chapter is to provide a description at the macroscopic scale of the early moments of the flow, namely the \emph{slumping regime}, with respect to the relevant dimensionless parameters. For this, we combine a total of 288 runs from three different lock-release devices and from two-fluids numerical simulations, which allow us to cover a large range of particle types (size and density), volume fractions, bottom slopes and geometries.
		%
		By tracking the front propagation through time, we extract the dimensionless slumping velocity $\mathcal{F}r$ and dimensionless characteristic slumping duration $\tau$, on which we base our description.
		%
		Our results show that the slumping velocity increases with the bottom slope, but decreases with the particle volume fraction when the latter exceeds a critical value. However, it remains independent of particle settling processes, which on the other hand affects the slumping duration. Hence, above a critical threshold, $\tau$ decreases as the ratio between the settling velocity and characteristic current velocity increases. For all these regimes, we derive scalings and energetic balances that reproduce the observed trends. The latter comparison confirms the role of initial energy transfer from the initial state towards the slumping phase on the resulting dynamics. This initial process and its characterisation remain crucial to prescribe relevant initial conditions for large-scale predictive modelling.

	\end{abstract}

	\vspace{0.3cm}
	{\it \color{red}REMINDER: Word limit of about 10,000 per article, plus 7 figures.}\\
	\vspace{0.3cm}
]

\section{Introduction}
\label{sec:intro}

\subsection{Finite volume density currents}
\label{sec:intro_lockrelease}

Density currents have been widely considered at the laboratory scale for their ubiquitous occurrence in geophysical flows \citep{Hopfinger1983,Simpson1999,Dufek2016}. As they are mostly associated with unsteady and non-uniform dynamics in nature, the lock-release device appeared to be a popular canonical configuration to mimic their behaviour at the laboratory scale \citep{Hacker1996,Shin2004,Nogueira2014}. In particular, studies often focused on characterising the front dynamics and the front shape as a macroscopic description of the underlying complex dynamics \citep{Rottman1983,Marino2005,Hogg2006}.

The basic approach to describe a lock-release evolution is built upon characteristics solutions of the hyperbolic shallow water equations for an inviscid dense liquid of density $\rho_{\text c}$ slumping on a horizontal bottom under an inertialess and weightless ambient fluid. In particular, this allows predicting the front velocity $U_{\text{c}}$ only based on the initial condition of the released mass. At the early stages of the slumping phase, it is found to be simply related to the only velocity scale of the system as
\begin{equation}
	U_{\text{c}}\propto \sqrt{gH_0},
\end{equation}
with $g$ gravitational acceleration and $H_0$ the initial height of the fluid volume being released. The method of characteristics suggests a coefficient of proportionality of exactly $2$ between velocities $U_{\text{c}}$ and $\sqrt{g H_0}$. An extension towards a weighted ambient fluid of density $\rho_{\text a}$ could lead to
\begin{equation}
	\label{eq:solUc}
	U_{\text{c}}=F\sqrt{\tilde{g}H_0},\quad\mbox{with}\quad \tilde{g}=\frac{\rho_{\text c}-\rho_{\text a}}{\rho_{\text c}}g,
\end{equation}
However, the coefficient of proportionality $F$, similar to a Froude number, now requires slightly more attention as it is clearly found smaller than $2$ in experiments or simulations. The main reason is that most density currents are characterized by a small density ratio between current and ambient, making the assumption of inertialess ambient fluid unlikely. A dynamic front condition is now required, and is often inspired from the steady front configuration derived by \citet{Benjamin1968}, now also linking the ambient height $H_{\text a}$ to the front velocity. This front relation includes two unknowns, a front velocity and a front height, and is often used as a front boundary condition in shallow water numerical solver for gravity currents \cite[e.g.][]{Ungarish2009}. Then, past studies managed to link the front velocity to the current height close to the front. However, the link between the front velocity and the initial conditions then becomes much more difficult to predict.
Models have been proposed to circumvent this issue, leading for instance to \citep{Huppert1980}
\begin{equation}
	\label{eq:Froude_Huppert1980}
	F = 0.5 \sqrt{\frac{\rho_{\text c}}{\rho_{\text a}}}\left(\frac{H_{\text a}}{H_0}\right)^{1/3}.
\end{equation}
%
The latter is relevant for not too deep ambient $H_{\text a}$ compared to the reservoir height $H_0$, which is actually the case in most experiments. Many configurations correspond to full-depth release ($H_{\text a}=H_0$), which leads to $F = 0.5(\rho_{\text c}/\rho_{\text a})^{1/2}$. Even if several experiments and numerical simulations confirm such a result, the exact value of $F$ remains still unclear and probably dependent on several physical processes. In particular, these relationships during the slumping regime are based on assuming negligible dissipation, which shall be violated in some cases, for instance in the presence of settling particles or significant mixing with the ambient at the miscible interface.

For a finite volume lock-release device on a mild enough slope, the slumping regime necessarily ends when the rarefaction wave associated with finite volume reaches the current front \citep[e.g.][]{Rottman1983, Hogg2006}. This leads to a new regime usually referred to as the inertial regime in the literature. This is observed on the front evolution as $U_{\text{c}}$ is no longer time independent but evolves as $t^{-1/3}$.
A simple approach can be useful to provide a rapid estimation of the position $X_{\text t}$ or time $T_{\text t}$ at which this transition occurs. For that purpose, it is worth mentioning that such a transition occurs as a rarefaction wave evolves faster than the front. Then, assuming the front velocity $U_{\text{c}}$ to be different from the rarefaction wave velocity at the interface $\sqrt{\tilde{g}H_0}$, which is assumed to be constant along its entire propagation path for simplicity, this leads to a transition length scale $X_{\text t}$ written as
\begin{equation}
	\label{eq:Xt_inertial}
	\displaystyle X_{\text t}\propto  \frac{2L_0 F}{1 - F}.
\end{equation}
Note that here $2$ comes from the path of propagation of the wave: emanating from the gate, it propagates backwards to the end of the reservoir, reflects on the wall and then propagates towards the front.
During the slumping regime, the previous estimation of $F$ \eqref{eq:Froude_Huppert1980} for full depth release $H_0=H_{\text a}$ leads to $X_{\text t}\propto 2L_0/[2(\rho_{\text a}/\rho_{\text c})^{1/2}-1]$. Note that, although more advanced methods based on single-layer shallow-water models predict a similar result, $X_{\text t} = 2 L_0 F (1 + F/2)^{1/2}$ \citep{Hogg2006}, full depth-release experiments have found much larger transitional positions, i.e $X_{\text t} \approx 10 L_{0}$~\citep{Rottman1983}.

Other mechanisms can modify the dynamics of the slumping regime which could lead to other transitions of the front velocity at different positions $X_{\text t}$ or times $T_{\text t}$, prior to the influence of the rarefaction wave. This is basically part of the topic of the present chapter.

For a gravity current on a horizontal bottom, a classic transition discussed in the literature is the one dominated by viscous dissipation. The flow eventually reaches a fully viscous regime \citep{Huppert1980}. This transition typically holds for fully laminar flow. A similar regime can be induced by inertial friction, for which the full derivation remains less obvious \citep[see for instance][for the case of a constant-influx current]{Hogg2001}. In any case, such transition can be seen as the time or position at which the current is strongly affected by dissipation, which actually encompasses entrainment for miscible gravity currents. The slumping regime then ends when dissipation induced by bottom friction and/or entrainment at the miscible interface, overcomes the current dynamics.
%
Again using a simple idealised approach, the latter transition can be estimated from a dynamic condition balancing an estimation of the shear stress $\rho_{\text c} C_d U_{\text{c}}^2$ ($C_d$ being some drag coefficient induced by friction at the bottom and/or by entrainment at the interface) with inertia over the current height $h_{\text c}$, $\rho_{\text c} U_{\text{c}}^2 h_{\text c} /X_t$. In the latter expression, $h_{\text c}$ can have different scaling. In the case of a clearly detached current head, the constant height scales with the initial height $H_0$. When assuming that the current remains attached to the initial column at transition \citep[similarly to][but for inertial drag instead of viscous dissipation]{Bougouin2022}, mass conservation between the initial column and a simplified box geometry implies $h_{\text c} = H_0 L_0/X_t$.
%
Both cases respectively lead to
\begin{equation}
	\label{eq:Xt_drag}
	\displaystyle X_{\text t}= {\frac{H_0}{C_d}}\quad \mbox{or} \quad X_{\text t}=  \sqrt{\frac{H_0 L_0}{C_d}}.
\end{equation}

To finish with, constant bottom slope $\alpha$, often relevant to characterize the main process of natural flow over topography, adds complexity to this front analysis. First, note that, when considering an inclined bottom, the reduced gravity and therefore the characteristic velocity \eqref{eq:solUc} shall be updated with a $\cos{\alpha}$ term accounting for the reduced effect of the initial column pressure normal to the inclined bottom. Second, including a bottom slope also adds a constant forcing along the slope suggesting a constant acceleration of the flow, at least until balanced by friction at both the interface and the bottom. Hence, it could affect the existence of the aforementioned slumping constant-velocity regime.
%
Yet, constant velocities associated with a slumping regime are clearly observed at early stages in experiments and simulations prior to such balance \citep{Birman2007, Gadal2023}. This is due to the relatively long time or length scales required for slope-induced acceleration to overcome the flow induced by the initial pressure gradient. A way to highlight that is a simple balance between slope and pressure gradient, resulting in
\begin{equation}
	\label{eq:Xt_slope}
	X_{\text t}=H_0\tan^{-1}{(\alpha)}.
\end{equation}
The latter can be seen as a transition length scale after which the slope starts driving the current until reaching the next equilibrium between slope and friction.

%%%%%%Figure 1
\begin{figure*}[ht!]
	\centering
	\includegraphics{figure1.pdf}
	\caption{\textbf{Experimental set-ups used in this study.} (a--c) Sketches of the three experimental set-ups. Below are shown snapshots of corresponding experimental images: (d--h) Set-up 1, $\alpha = 7^\circ$, $\phi \sim 3~\%$, ${S}t=0.027$; (i--k) Set-up 2, $\alpha = 45.4^\circ$, $\phi \sim 1.5~\%$, ${S}t=0.06 $; (l--p) Set-up 3, $\alpha = 0^\circ$, $\phi \sim 55\%$, ${S}t \equiv 0$; (q--v) SedFoam, $\alpha = 45.4^\circ$, $\phi \sim 1.7~\%$, $\mathcal{S}t=0.06$.}
	\label{fig:fig1}
\end{figure*}
%%%%%%%%

\subsection{Particle-Laden Gravity Currents}
\label{sec:intro_PLGC}

Particle-Laden Gravity Currents (PLGCs) refer to as gravity currents, or density currents, are driven by a mass difference between a heavy fluid-particle mixture and a lighter ambient liquid \citep[e.g.][]{Hopfinger1983,Middleton1993,Kneller2000,Meiburg2010,wells2021}. Different subclasses of PLGCs can be distinguished depending on the relative density of the different phases constituting the system; i.e. the particle phase, the carrier fluid phase and the ambient fluid phase. In the following, we will focus on two of them, characterized by only two distinct densities. At one end, the carrier phase and ambient phase are the same fluid and particles are heavier. This is the most commonly studied subclass of PLGCs, usually referred to as turbidity currents. At the other end, PLGCs can also be induced by a neutrally-buoyant fluid-particle mixture flowing into a lighter ambient fluid.

A major difference between PLGCs and compositional currents is therefore the existence of typical length and time scales associated with the particle dynamics. Nevertheless, the prediction of these finite scales is difficult as it shall strongly depend on the interaction between particle dynamics and the local turbulence of the carrying fluid and/or ambient fluid at the upper interface which modifies entrainment and associated dissipation. In turn, this can modify the macroscopic behaviour of the current, for instance, the dynamics of the front. Accordingly, these dynamics shall be controlled by several dimensionless parameters that can lead to a wide variety of flow regimes. In particular, beyond the slope $\alpha$, the dimensionless density difference between the current and the ambient $\mathcal{A}t$, the ratio between available inertia and dissipation, say a Reynolds number $\mathcal{R}e$ as usually introduced as a control parameter for any density currents, the initial solid fraction of particles $\phi$ and a dimensionless settling velocity $S=V_{\text s}/U_{\text{c}}$ with $V_{\text s}$ the settling velocity of particles into the carrying fluid phase, have now to be considered as potentially affecting flow regimes. It shall be noted that in some specific situations, some of these dimensionless parameters could be linked (for instance $\phi$ and $\mathcal{A}t$ for heavy particles in a single-phase liquid for the carrying and ambient fluids).

Beyond the transition from the slumping regime to other regimes discussed in the previous section (still relevant for PLGCs although parametrisation of the processes involves more complex mechanisms induced by the fluid-particle interaction), here, a new transition towards the deposition of the particle can emerge. In particular, based on a kinematic approach balancing time scales related to current propagation and particle settling, leads to
%
\begin{equation}
	\label{eq:Xt_settling}
	X_{\text t} \propto H_0\frac{U_{\text{c}}}{V_{\text s}}.
\end{equation}
%
Note that such transition shall disappear for the neutrally-buoyant situation of PLGCs.
However, it has been observed that mixing at the interface can modify the local concentration of particles at the front, resulting in behaviour similar to that of settling~\cite{Schneider2023}. Again, such an approach only gives a scaling of transition, but the influence of the particle dynamics prior to such transition remains unclear. Overall, the value of $F$ in \eqref{eq:solUc} during the slumping inertial regime remains to clarify varying the influence of the mechanisms discussed so far.

Even if the dynamics of PLGCs is associated with several complex processes, a general consequence is observed in the front dynamics, often referred to as the Froude number, and the front shape. Then prediction of the different flow regimes associated with PLGCs shall be somewhat included in these observables. After presenting the experimental and numerical methods used to investigate PLGCs dynamics in section~\ref{sec:methods}, we introduce the main dimensionless quantities used and macroscopic observations in section~\ref{sec:current_description}. We discuss our results in section~\ref{sec:results} and conclude in section~\ref{sec:conclusion}.

\section{Methods}
\label{sec:methods}

In order to characterize the slumping regime, and its transition to other regimes, of PLGCs induced by the finite-volume release of a suspension initially at ``rest'', 3 lock-release devices have been used to cover a wide range of control parameters. These experimental devices are supplemented with two-fluid numerical simulations to confirm the relevance of such a flow description to characterize the dynamics of PLGCs.

\subsection{Experimental set-ups}

All set-ups are based on the same typical lock-release design (see figure~\ref{fig:fig1}). The tank has a width $W$, a height $H$ and a length $L_0 + L_1$ along the $x$ direction. It is split into two parts by a sluice gate at a distance $L_0$ from the left wall. %More specifically, a tank of width $W$, height $H$ and length $L_0+L_1$ along the streamwise direction $x$ is split in two along this direction by a sluice gate at a distance $L_0$ from one end. 
On the $L_0$ side, a reservoir is filled up to a height $H_0$ with a fluid/particle mixture having a density $\rho_{\text{m}} = \phi \rho_{\text{p}} + (1 - \phi)\rho_{\text{f}}$, with $\rho_{\text{p}}$ and $\rho_{\text{f}}$ the particle and fluid densities, respectively. On the other side of height $H_{\text a}$ is the ambient fluid of density $\rho_{\text{a}}$, with typically $\rho_{\text{a}} < \rho_{\text{m}}$. At $t = 0$, the sluice gate is lifted up manually to initiate the gravity current.

@@

The first set-up (figure~\ref{fig:fig1}(a)) has $L_0+L_1=150~{\text cm}$ with $L_0 = 10~\textrm{cm}$, $W = 10~\textrm{cm}$ and $H=49~\textrm{cm}$. The upper boundary is a free surface and the tank can be inclined at an angle $\alpha\in[0^\circ,7.5^\circ]$ with respect to the horizontal as shown in the 2D sketch of figure~\ref{fig:fig1}(a). For inclined configurations, $H_{\text a}$ then increases along the flume, and the reservoir is filled such as both heights at the gate, reservoir one and ambient one, are the same (see figure~\ref{fig:fig1}(a)). The suspension in the reservoir is made by strongly stirring a known mass of particles within the whole water column using a mechanical mixer, which is stopped just before opening the gate. $H_{0}$ is defined as the average suspension height inside the lock, equal to $20~\textrm{cm}$.
%
The current evolution is followed by an sCMOS pco.edge$^\copyright$ camera at 120 Hz, using a LED backlight as a light source. Additional details on the set-up and the image processing can be found in \citet{Gadal2023}.

\paragraph{Set-up 2}
The second set-up (figure~\ref{fig:fig1}(b)) has $L_0+L_1=3.8~{\text m}$ with $L_0 = 30~\textrm{cm}$ for runs with PMMA particles and $L_0 = 10~\textrm{cm}$ for runs with silica sand or hydrogel particles (see section~\ref{sec:datasets} for particles specifications). The overall flume width and depth are $W= 10~\textrm{cm}$ and $H= 50~\textrm{cm}$. The upper boundary in the flume is a rigid lid, and the only free surface is in the lock region. The ambient fluid is tap water at room temperature and $H_{\text a}=50~\textrm{cm}$. The flume can be inclined at an angle $\alpha \in [0^\circ,45.4^\circ]$ as shown in figure~\ref{fig:fig1}(b). Here, $H_0$ is defined as the average suspension depth perpendicular to the flume floor inside the lock at the gate opening time. LED plates illuminate the flume from above. The flow dynamic is recorded with a Phantom$^\copyright$ Miro C110 camera at 50 Hz.

@@

The third set-up (figure~\ref{fig:fig1}(c)) has $L_0+L_1=150~{\text cm}$ with $L_0 = 20~\textrm{cm}$, $W= 10~\textrm{cm}$ and $H = 10~\textrm{cm}$. The upper boundary is a free surface and the tank remains horizontal ($\alpha = 0^\circ$), as shown in the 2D sketch of figure~\ref{fig:fig1}(c). Here, the suspension and ambient heights are equal, i.e. $H_{a}=H_{0}$, and vary between 20~cm and 30~cm depending on experiments. The current evolution is followed by a Nikon$^\copyright$ D500 camera at 10 Hz and using a white LED panel as a light source. Additional details on the set-up and the image processing can be found in \citet{Schneider2023}.

\paragraph{Two-fluid model}

A two-fluid model, SedFoam \citep{chauchat2017}, is used to simulate lock-release configurations from set-ups 1 and 2. This model available for download at Zenodo \citep{bonamy2023} is implemented in the open-source computational fluid dynamics toolbox OpenFOAM and has been extensively validated for various sediment transport problems such as sheet-flow \citep{cheng2016,chauchat2022,mathieu2021,mathieu2022}, scour \citep{nagel2020,tsai2022}, ripples \citep{salimi2021a,salimi2021b}, avalanches \citep{montella2021} and granular collapse \citep{montella2023}. The simulations presented herein are 2D vertical and turbulence-averaged. The turbulence model is a $k-\epsilon$ model with additional two-phase flow terms for density stratification and turbulent drag dissipation. The closure for the particle-particle interactions is a frictional-collisional model based on a Coulomb rheology for the frictional part and the kinetic theory of granular flows for the collisional part. The two contributions are simply added as described in \cite{chauchat2017}. The drag force is modelled using the \cite{schiller1933} closure multiplied by a hindrance function of \cite{richardson1954}. In order to describe the influence of the slope, the total pressure is solved and the gravity vector is rotated while the mesh is kept aligned with the $x$ and $y$ axis.

\subsection{Datasets}
\label{sec:datasets}

269 experiments have been performed on the 3 experimental set-ups described in the previous section. They are complemented by 19 numerical simulations using SedFoam. Altogether, they are classified into 4 different datasets corresponding to the 3 experimental set-ups and the numerical modelling respectively. These 4 datasets are described below.

@@

The first dataset is gathered using set-up 1. A first series of experiments is carried out with silica sand grains with $\rho_p=2650\textrm{kg/m}^{3}$ and $d \sim 120~\mu\textrm{m}$ while varying the bottom slope from $0^\circ$ to $7^\circ$. Moreover, for $\alpha = 7^\circ$, the settling velocity is varied by using glass beads (Silibeads$^\copyright$) with $\rho_p=2500\textrm{kg/m}^{3}$ and mean diameter ranging $d$ from $60~\mu\textrm{m}$ to $250~\mu\textrm{m}$, corresponding to $V_{\text s} \in [0.3, 3.2]~\textrm{cm}~\textrm{s}^{-1}$. In most series of experiments, the initial volume fraction $\phi$  is systematically varied in the range $[0.003, 0.15]$.
%
A few experiments with saline water (homogeneous gravity currents) performed at $\alpha=7^\circ$ are also included in the dataset, however not discussed in this chapter.

@@

The second dataset is gathered using set-up 2. Three different types of particles have been used in order to vary $d$, $\rho_{\text p}$ and as a direct consequence $V_{\text s}$ and $\mathcal{S}t$. A first series of experiments is carried out with the same silica sand grains as in dataset 1, with the same stirring procedure and varying the bottom slope from $7^\circ$ to $15^\circ$ and $\phi$ from 0.01 to 0.1.
%
A second series of experiments is carried out with PMMA particles with $\rho_{\text p}=1190~\textrm{kg/m}^{3}$ and diameter $d\sim 250~\mu\textrm{m}$. Part of the runs had a manual stirring with a rod and $H_0=H_{\text a}$, while others deal with a release of a deposited layer of particles and no mixing leading to $H_0<H_{\text a}$. Here, the bottom slope is varied between $0^\circ$ and $45.4^\circ$, and the initial volume fraction is kept nearly constant, i.e. $\phi \in [0.01, 0.02]$.
%
A third series of experiments is carried out with hydrogel particles of densities $\rho_{\text p}\in \{1003, 1045\}~\textrm{kg/m}^{3}$ and $d\in \{0.3, 1.6\}~\textrm{cm}$. For most of them, the flume was inclined to $45.4^\circ$ and experiments began with particles deposited inside the lock (no stirring, $H_0<H_{\text a}$). In these cases, hydrogel physical properties, among which deformability, make volume fraction within the deposited layer reach values up to nearly 1 (full packing), way above the classical 0.65 for solid spherical monodisperse granular materials. These initially deposited runs are highlighted in each figure by red-edged symbols. Additionally, a couple of hydrogel runs have been made at a lower slope angle and were released from a stirred configuration.

@@

The third dataset is gathered using setup 3. The experiments are performed using a neutrally-buoyant fluid-particle mixture made of polystyrene beads of typical diameter $d \sim 1~\textrm{mm}$ with density $\rho_{\text p} \sim 1038.5~\textrm{kg/m}^{3}$. Saltwater is added to the lock for density matching. The density difference between the lock and the ambient fluid is kept nearly constant for each experiment, equal to $25 \textrm{kg}~\textrm{m}^{-3}$. Only the initial volume fraction is varied in the range $[0.03$  $0.55]$.

@@

This dataset has been obtained using the two-fluid model described above. The domain size is $L_0+L_1=3.8$~m long and $W=0.5$~m height and the typical grid resolution is $\Delta x=1.25$~mm in the streamwise direction and it varies between $\Delta y \in [0.25 ; 3.5]$ mm in the wall-normal direction with a geometrical progression. The time step is adaptative following criteria on the Courant number, $C_{FL}=\max(U) \Delta t / \Delta x < 0.3$ for the phase velocities as well as for the relative velocity between the phases (the maximum allowed time-step during the simulation is set to $\max(\Delta t) = 10^{-3}$~s). The lock is $L_0=0.3$~m wide and $H_0=0.5$~m high, corresponding to the domain dimensions. The numerical schemes are second order in time and space: backward for time, a blend of centred and upwind schemes for advection (limitedLinear in OpenFOAM) and a centred scheme for diffusion. The initial condition is fluid and particle phase at rest ($u_f=u_p=0$) and the initial volume fraction of particles $\phi$ is set to a constant value in the region $x\in[-0.5 ; 0]$ m and $y\in[0;0.5]m$. At $t=0$, the governing equations are numerically integrated until $t_{end}=80$s maximum. The front position is determined as the most downstream position of the isocontour of volume fraction $\phi =10^{-3}$ at a frequency of 10Hz. This criterion has been demonstrated to be robust for all the simulations presented in this work. The particle properties are $e=0.8$ (restitution coefficient for binary collisions) and $\mu_{\text s}=0.53$ (macroscopic friction coefficient). Two sets of simulations have been performed. One mimics silica sand beads in water $\rho_{\text p}/\rho_{\text f}=2.65$ with diameter $d_p=60~\mu$m and initial volume fractions $\phi\in[0.015, 0.1]$. The other one mimics PMMA particles in water $\rho_{\text p}/\rho_{\text f}=1.19$ with diameters $d \in [100, 650]~\mu$m and initial volume fractions $\phi \in [0.0144, 0.5]$. For both configurations, the slope angle $\alpha$ has been varied in the range $\alpha \in [0^\circ, 45^\circ]$.
